### 2023-09-20 proofs/main.tex
Elaborate on the algorithmic typing

\begin{algorithm}[Type Well-formedness]
  \hfill

@@

\begin{algorithm}[Type Normalization]
  \hfill

@@

\begin{algorithm}[Subtyping]
  \hfill

@@

Next, we define the specification of the type 
inference for our language.
The judgement $[[Γ ; Φ ⊢ v : iP]]$
means that under the type context $[[Γ]]$ and
the variable context $[[Φ]]$, for the value $[[v]]$,
type $[[iP]]$ is inferrable. The negative form of the judgement is symmetric.
The application inference judgement $[[ Γ ; Φ ⊢ iN ● args ⇒> iM ]]$ means that if
a head of type $[[iN]]$ is applied to list of values $[[args]]$,
then the resulting computation can be typed as $[[iM]]$.

@@

\begin{definition}[Declarative Type Inference]
  \hfill
  \begin{multicols}{2}
  \ottdefnDTNInf{}
  \ottdefnDTPInf{}
  \ottdefnDTSpinInf{}
  \end{multicols}
\end{definition}

@@

\subsection{Algorithmic Type Inference}
\begin{algorithm}
  \hfill
  \begin{multicols}{2}
  \ottdefnATNInf{}

@@

@@

\begin{algorithm}
  \hfill
  \begin{multicols}{2}
    \ottdefnSINGscEN{}

@@

>>>>

\begin{algorithm}[Type Well-formedness]
  \label{alg:wf}
  \hfill

@@

\begin{algorithm}[Type Normalization]
  \label{alg:type-nf}
  \hfill

@@

\begin{algorithm}[Subtyping]
  \label{alg:subtyping}
  \hfill

@@

Next, we define the specification of the type 
inference for our language. First, we introduce 
variable context specifying the types of the variables 
in the scope of the current rule.

@@

The specification is represented by an inference system of
three mutually recursive judgments:
positive inference $[[Γ ; Φ ⊢ v : iP]]$,
negative type inference $[[Γ ; Φ ⊢ c : iN]]$, and
application type inference $[[Γ ; Φ ⊢ iN ● args ⇒> iM ]]$.
In the premises, the inference rules also refer to 
the declarative subtyping (\cref{def:subDOne}),
type well-formedness (\cref{alg:wf}), 
and normalization (\cref{alg:alg:type-nf}) judgements.
\begin{itemize}
  \item $[[Γ ; Φ ⊢ v : iP]]$ (and symmetrically, $[[Γ ; Φ ⊢ c : iN]]$)
    means that under the type context $[[Γ]]$ and
    the variable context $[[Φ]]$, for the value $[[v]]$,
    type $[[iP]]$ is inferrable. It guarantees that 
    $[[v]]$ is well-formed in $[[Γ]]$ and $[[Φ]]$ in
    the standard sense.
  \item $[[ Γ ; Φ ⊢ iN ● args ⇒> iM ]]$ is the application type inference 
    judgement. It means that if a head of type $[[iN]]$ 
    is applied to list of values $[[args]]$, 
    then the resulting computation can be typed as $[[iM]]$.
\end{itemize}

\begin{definition}[Declarative Type Inference]
  \label{def:declarative-typing}
  \hfill
  \begin{multicols}{2}
  \ottdefnDTNInf{}
  \ottdefnDTPInf{}
  \ottdefnDTSpinInf{}
  \end{multicols}
\end{definition}

@@

Next, we present the type inference algorithm, 
which is sound and complete with respect to the declarative specification
(\cref{def:declarative-typing}).

\subsection{Algorithmic Type Inference}

Mirroring the declarative typing, 
the algorithm is represented by an inference system of three mutually recursive
judgments:
\begin{itemize}
  \item $[[Γ ; Φ ⊨ v : iP]]$ and  $[[Γ ; Φ ⊨ c : iN]]$
    are the algorithmic versions of $[[Γ ; Φ ⊢ v : iP]]$ and $[[Γ ; Φ ⊢ c : iN]]$.
    In contrast with the declarative counterparts, they are deterministic,
    and  guarantee that the inferred type is normalized. 
  \item $[[Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC]]$
    is the algorithmization of $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$.
    Notice that $[[uN]]$ contains algorithmic variables, 
    which are specified by the context $[[Θ1]]$.
    Moreover, the inferred type $[[uM]]$ is also algorithmic,
    and can have several non-equivalent instantiations. To accommodate that, 
    the algorithm also returns $[[Θ2]]$ and $[[SC]]$ specifying 
    the variables used in $[[uM]]$: $[[Θ2]]$ defines the contexts
    in which the variables must be instantiated, and $[[SC]]$
    imposes restrictions on the variables. 
\end{itemize}
As subroutines, the algorithm calls
subtyping (\cref{alg:subtyping}),
type well-formedness (\cref{alg:wf}),
constraint merge (\cref{sec:constraint-merge}),
normalization (\cref{alg:type-nf}),
and constraint singularity which will be defined later in \cref{todo}.
It also relies on basic set operations and the ability to 
deterministically choose fresh variables.

\begin{algorithm}
  \hfill
  \begin{multicols}{2}
  \ottdefnATNInf{}

@@

Let us discuss the inference rules of the algorithm:
\begin{itemize}
  \item \ruleref{\ottdruleATVarLabel} 
    infers the type of a variable by looking it up in the context
    and normalizing the result.
  \item \ruleref{\ottdruleATThunkLabel} and \ruleref{\ottdruleATReturnLabel}
    are similar to the declarative rules: they make a recursive call
    to type the body of the thunk or the return expression and
    put the shift on top of the result.
  \item \ruleref{\ottdruleATPAnnotLabel} and \ruleref{\ottdruleATNAnnotLabel}
    are symmetric. They make a recursive call to infer the type of the annotated
    expression, check that the inferred type is a subtype of the annotation,
    and return the normalized annotation.
  \item \ruleref{\ottdruleATtLamLabel} infers the type of a lambda-abstraction.
    It makes a recursive call to infer the type of the body in the extended context,
    and returns the corresponding arrow type. Notice that the algorithm also
    normalizes the result, which is because the annotation type $[[iP]]$
    is allowed to be non-normalized.
  \item \ruleref{\ottdruleATTLamLabel} infers the type of a big lambda.
    Similarly to the previous case, it makes a recursive call to infer the type
    of the body in the extended \emph{type} context. 
    After that, it returns the corresponding universal type. 
    It is also required to normalize the result, because, 
    for instance, $[[α⁺]]$ might not occur in the body of the lambda,
    in which case the $[[∀]]$ must be removed. 
  \item \ruleref{\ottdruleATVarLetLabel} is defined in a standard way:
    it makes a recursive call to infer the type of the bound value,
    and then returns the type of the body in the extended context.
  \item \ruleref{\ottdruleATAppLetAnnLabel}
    is interpreted as follows.
    First, it infers the type of the head of the application,
    ensuring that it is a thunked computation $[[↓iM]]$;
    after that, it makes a recursive call
    to the application inference procedure,
    which returns the algorithmic type, whose
    instantiation to a declarative type must be associated with the bound variable 
    $[[x]]$; then premise $[[Γ; Θ ⊨ ↑uQ ≤ ↑iP ⫤ SC2]]$
    together with $[[Θ ⊢ SC1 & SC2 = SC]]$
    check whether the instantiation to the annotated type $[[iP]]$ is possible,
    and if it is, the algorithm infers the type of the body in the extended context,
    and returns it as the result. 
  \item \ruleref{\ottdruleATAppLetAnnLabel}
    works similarly to \ruleref{\ottdruleATAppLetAnnLabel},
    However, since there is no annotation, 
    instead of checking the instantiation to it, 
    the algorithm checks that the inferred type
    $[[↑uQ]]$ is unique.
    It is the case if all the algorithmic variables of 
    $[[↑uQ]]$ are sufficiently restricted by $[[SC]]$,
    which is checked by the combination of
    $[[uv uQ = dom(SC)]]$ and $[[SC singular with uσ]]$.
    Together, these two premises guarantee that the only 
    possible instantiation of $[[↑uQ]]$ is $[[ [uσ]uQ ]]$.
  \item \ruleref{\ottdruleATUnpackLabel}
    works in the expected way. First, it infers the 
    existential type $[[∃nas.iP]]$ of the value being unpacked,
    and since the type is guaranteed to be normalized, binds 
    the quantified variables with $[[nas]]$.
    Then it infers the type of the body in the appropriately extended context,
    and checks that the inferred type does not depend on $[[nas]]$
    by checking well-formedness $[[Γ ⊢ iN]]$.
\end{itemize}

Finally, let us discuss the algorithmic rules of the application inference:
\begin{itemize}
  \item \ruleref{\ottdruleATEmptyAppLabel}
    is the base case. If the list of arguments is empty, 
    the inferred type is the type of the head,
    and the algorithm returns it after normalizing.
  \item \ruleref{\ottdruleATArrowAppLabel}
    is the main rule of the algorithmic application inference.
    If the head has an arrow type $[[uQ → uN]]$,
    we find $[[SC1]]$---the minimal constraint ensuring that 
    $[[uQ]]$ is a supertype of the first argument's type.
    Then we make a recursive call applying $[[uN]]$ to the rest of the arguments,
    and merge the resulting constraint with $[[SC1]]$
  \item \ruleref{\ottdruleATForallAppLabel},
    analogously to the declarative case,
    is the rule ensuring the implicit elimination of the universal quantifiers. 
    This is the place where the algorithmic variables are generated.
    The algorithm simply replaces the quantified variables 
    $[[pas]]$ with fresh algorithmic variables $[[puas]]$,
    and makes a recursive call in the extended context. 
\end{itemize}

The correctness of the algorithm consists of its soundness and 
completeness, which is by mutual
induction in \cref{lemma:typing-soundness,lemma:typing-completeness}.
The simplified result is the following.
\begin{theorempreview}
  \hfill
  \begin{itemize}
    \item [$-$] $[[Γ; Φ ⊨ c : iN]]$ implies $[[Γ; Φ ⊢ c : iN]]$, 
      and $[[Γ; Φ ⊢ c : iN]]$ implies $[[Γ; Φ ⊨ c : nf(iN)]]$;
    \item [$+$] $[[Γ; Φ ⊨ v : iP]]$ implies $[[Γ; Φ ⊢ v : iP]]$, 
      and $[[Γ; Φ ⊢ v : iP]]$ implies $[[Γ; Φ ⊨ v : nf(iP)]]$.
  \end{itemize}
\end{theorempreview}

@@

The singularity algorithm used in  \ruleref{\ottdruleATAppLetAnnLabel}
of the algorithmic typing check wether the constraint $[[SC]]$
is uniquely defines a substitution satisfying it, and if it does
returns such a substitution as the result.
To do that, we define a partial function $[[SC singular with uσ]]$,
taking a subtyping constraint $[[SC]]$ and returning a substitution 
$[[uσ]]$---the only possible solution of $[[SC]]$.

First, we define the notion of singularity on constraint entries. 
$[[scE singular with iP]]$ and $[[scE singular with iN]]$
are considered partial functions taking a constraint entry $[[scE]]$
and returning the type satisfying $[[scE]]$ if such a type is unique.

\begin{algorithm}[Singular Constraint Entry]
  \hfill
  \begin{multicols}{2}
    \ottdefnSINGscEN{}

@@

\begin{itemize}
  \item \ruleref{\ottdruleSINGNEqLabel} and \ruleref{\ottdruleSINGPEqLabel} 
    are symmetric. If the constraint entry says that a variable must be equivalent to 
    a type $T$, then it is evidently singular, and the only (up-to-equivalence) type
    instantiating this variable could be $T$. This way, we return its normal form. 
  \item \ruleref{\ottdruleSINGSupVarLabel}
    implies that the only (normalized) solution of $[[pua :≥ ∃nas.pa]]$ is 
    $[[pa]]$ (it will be shown in \cref{lemma:var-subt}).
  \item \ruleref{\ottdruleSINGSupShiftLabel}
    is perhaps the least obvious rule.
    In type $[[∃nas.↓iN]]$, if $[[iN]]$ is anything different (non-equivalent) to 
    $[[nai ∊ {nas}]]$, there are at least 
    one proper supertype of $[[∃nas.↓iN]]$, since $[[iN]]$ can be abstracted over
    by an existential quantifier. Otherwise, any supertype of $[[∃nas.↓nai]]$
    is equivalent to it, and thus, the solution is unique. 
\end{itemize}

Next, we extrapolate the singularity function on constraints---sets of constraint entries. 
We require $[[SC]]$ to be a set of singular constraints, and the resulting substitution
sends each variable from $[[dom(SC)]]$ to the unique type satisfying the corresponding constraint.

\begin{algorithm}
  $[[SC singular with uσ]]$
  means that 
  \begin{enumerate}
    \item for any positive $[[scE ∊ SC]]$,
      there exists $[[iP]]$ such that $[[scE singular with iP]]$, 
      and for any negative $[[scE ∊ SC]]$,
      there exists $[[iN]]$ such that $[[scE singular with iN]]$;
    \item $[[uσ]]$ is defined as follows:
      $$
      [[ [uσ]β̂⁺ ]]  = 
          \begin{cases}
              [[ iP ]] & \text{if there is } [[scE]] \in [[dom(SC)]] \text{ restricting } [[β̂⁺]] 
                         \text{ and } [[scE singular with iP]] \\
              [[ β̂⁺ ]] & \text{otherwise}  \\
          \end{cases}
      $$
      $$
      [[ [uσ]β̂⁻ ]]  = 
          \begin{cases}
              [[ iN ]] & \text{if there is } [[scE]] \in [[dom(SC)]] \text{ restricting } [[β̂⁻]] 
                         \text{ and } [[scE singular with iN]]\\
              [[ β̂⁻ ]] & \text{otherwise}  \\
          \end{cases}
      $$
  \end{enumerate} 
\end{algorithm}

The correctness of the singularity algorithm is formulated as follows:
\begin{theorempreview}
  Suppose that $[[SC]]$ is a subtyping constraint.
  Then $[[SC singular with uσ]]$ holds if and only if 
  $[[uσ]]$ is the only (up-to-equivalence on $[[dom(SC)]]$) 
  normalized substitution satisfying $[[SC]]$.
\end{theorempreview}
